---
title: "COVID-19"
author: "Katie Forthman"
date: "`r date()`"
output: html_document
---

# ---- Load packages ----

```{r}
library(stringr)
library(corrplot)
```

# ---- Load neighborhood data ----

```{r}

load("Data/county_factors.rda")

```


# ---- Load  and format covid data ----

```{r}
data.path <- "Data/COVID-19/csse_covid_19_data/csse_covid_19_time_series/"

# Read in the data
US.deaths <- read.csv(
  paste0(data.path, "time_series_covid19_deaths_US.csv"), 
  header = T, stringsAsFactors = F)
US.cases <- read.csv(
  paste0(data.path, "time_series_covid19_confirmed_US.csv"), 
  header = T, stringsAsFactors = F)

# Read in the header seprately.
US.cases.head <- read.csv(
  paste0(data.path, "time_series_covid19_confirmed_US.csv"), 
  header = F, stringsAsFactors = F)[1,]
US.deaths.head <- read.csv(
  paste0(data.path, "time_series_covid19_deaths_US.csv"), 
  header = F, stringsAsFactors = F)[1,]

# Correct the dates in the header to be more useable as
# column names.
proper_date <- function(dates){
  dates <- sapply(dates, strsplit, split = "/")
  dates <- lapply(dates, str_pad, width = 2, side = "left", pad = "0")
  dates <- lapply(dates, paste, collapse = "_")
  dates <- unlist(dates)
  
  return(dates)
}

dates.cases <- proper_date(US.cases.head[-c(1:11)])
dates.deaths <- proper_date(US.deaths.head[-c(1:12)])

names(US.cases) <- c(US.cases.head[1,1:11], dates.cases)
names(US.deaths) <- c(US.deaths.head[1,1:12], dates.deaths)

if(sum(US.cases$UID != US.deaths$UID) > 0){warning("COVID data rows do not match!")}
US.cases$Population <- US.deaths$Population
US.cases <- US.cases[,c(1:11, ncol(US.cases), 12:(ncol(US.cases)-1))]
```


## Split the dataset into the data and the info for usability.
```{r}
US.cases.info <- as.matrix(US.cases[,1:12])
US.cases.data <- as.matrix(US.cases[,-c(2:12)])
US.deaths.info <- as.matrix(US.deaths[,1:12])
US.deaths.data <- as.matrix(US.deaths[,-c(2:12)])

rownames(US.cases.info) <- US.cases.info[,1]
US.cases.info <- US.cases.info[,-1]
rownames(US.cases.data) <- US.cases.data[,1]
US.cases.data <- US.cases.data[,-1]
rownames(US.deaths.info) <- US.deaths.info[,1]
US.deaths.info <- US.deaths.info[,-1]
rownames(US.deaths.data) <- US.deaths.data[,1]
US.deaths.data <- US.deaths.data[,-1]


ndays.cases <- ncol(US.cases.data)
ndays.deaths <- ncol(US.deaths.data)

nobs <- nrow(US.cases.data)
```


# ---- The Curve ----

```{r}
state.curve <- function(state, stat = c("cases", "deaths")){
  if(stat == "cases"){
    data <- US.cases.data[which(US.cases$Province_State == state),]
  }else if(stat == "deaths"){
    data <- US.deaths.data[which(US.deaths$Province_State == state),]
  }
  data.sum <- colSums(data)
  day.first.case <- min(which(data.sum > 0))
  n.days <- length(data.sum)
  barplot(data.sum[day.first.case:n.days], 
          main = paste0("Total COVID-19 ", stat," by date in ", state), 
          log = "y", las = 2, cex.axis = 1, cex.names = 0.5)
}

state.curve("New York", "cases")
state.curve("New York", "deaths")

state.curve("California", "cases")
state.curve("California", "deaths")

state.curve("Oklahoma", "cases")
state.curve("Oklahoma", "deaths")
```

```{r}
county.curve <- function(county, stat = c("cases", "deaths")){
  if(stat == "cases"){
    data <- US.cases.data[which(US.cases$Admin2 == county),]
  }else if(stat == "deaths"){
    data <- US.deaths.data[which(US.deaths$Admin2 == county),]
  }
  
  day.first.case <- min(which(data > 0))
  n.days <- length(data)
  
barplot(data[day.first.case:n.days], main = paste0("Total COVID-19 ", stat," by date in ", county), log = "y", las = 2, cex.axis = 1, cex.names = 0.5)
  
}

county.curve("Tulsa", "cases")
county.curve("Tulsa", "deaths")
```


# ---- Calculate some useful stats to compare with neighborhood data ----

```{r}
US.stats <- data.frame(UID = US.cases$UID)
```

## Average rise in cases per day
avg.rise.cases
```{r}
rise.cases <- matrix(ncol = ndays.cases - 1, nrow = nobs)
colnames(rise.cases) <- colnames(US.cases.data)[-1]
for(i in 1:ncol(rise.cases) + 1){
  rise <- US.cases.data[,i] - US.cases.data[,i-1]
  rise.cases[,i-1] <- rise
}

US.stats$avg.rise.cases <- apply(rise.cases, 1, mean)

rise.cases.total <- colSums(rise.cases)
barplot(rise.cases.total, main = "Rise in Cases of COVID-19 by Date in US", las = 2, cex.axis = 1, cex.names = 0.5)
```

## Average rise in deaths per day
avg.rise.deaths
```{r}
rise.deaths <- matrix(ncol = ndays.deaths - 1, nrow = nobs)
colnames(rise.deaths) <- colnames(US.deaths.data)[-1]
for(i in 1:ncol(rise.deaths) + 1){
  rise <- US.deaths.data[,i] - US.deaths.data[,i-1]
  rise.deaths[,i-1] <- rise
}

US.stats$avg.rise.deaths <- apply(rise.deaths, 1, mean)

rise.deaths.total <- colSums(rise.deaths)
barplot(rise.deaths.total, main = "Rise in Deaths of COVID-19 by Date in US", las = 2, cex.axis = 1, cex.names = 0.5)
```

## Total cases
total.cases
```{r}
US.stats$total.cases <- US.cases.data[,ndays.cases]
```

## Total cases per capita
```{r}
US.stats$total.cases.percap <- US.stats$total.cases / US.cases$Population
US.stats$total.cases.percap[US.cases$Population == 0] <- NA
hist(US.stats$total.cases.percap)
```

## Total deaths
total.deaths
```{r}
US.stats$total.deaths <- US.deaths.data[,ndays.deaths]
```

## Total deaths per capita
total.deaths.percap
```{r}
US.stats$total.deaths.percap <- US.stats$total.deaths / US.deaths$Population
US.stats$total.deaths.percap[US.deaths$Population == 0] <- NA

max(US.stats$total.deaths.percap,na.rm = T)
```

## Total deaths per case
total.deaths.percase
```{r}
# pos.case.ind <- US.stats$total.cases > 0
# US.stats$total.deaths.percase[pos.case.ind] <- US.stats$total.deaths[pos.case.ind] / US.stats$total.cases[pos.case.ind]
# US.stats$total.deaths.percase[!pos.case.ind] <- 0
US.stats$total.deaths.percase <- US.stats$total.deaths / US.stats$total.cases
US.stats$total.deaths.percase[US.stats$total.cases == 0] <- NA

US.stats[which(US.stats$total.deaths > US.stats$total.cases),]

```


# ---- Merge COVID data with Neighborhood data ----

```{r}
US.stats$ID <- str_pad(US.stats$UID, 8, "left", pad = "0")
US.stats$ID <- substr(US.stats$ID, 4, 8)

data.merge <- merge(US.stats, county_factors, by = "ID")
```


# ---- Plot the relationships ----

```{r}
data.cor <- cor(data.merge[,-c(1:2)], use = "complete.obs", method = "spearman")
corrplot.mixed(data.cor, upper = 'ellipse', lower = 'number', tl.pos = 'lt', tl.cex = 1, lower.col = "black", number.cex = 0.5)
```





